---
title: Batter's Swing Speed as Exogenous and Endogenous Variable.
       CSAS 2025 Data Challenge
author: "Vladimir Averin, TT Bulldogs"
date: "2025/01/14"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

<style type="text/css">
  body{
  font-size: 15pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo   =FALSE,      ## show or suppress the code
                      include=TRUE ,      ## show or suppress the output
                      message=FALSE,      ## omit messages generated by code
                      warning=FALSE,      ## omit warnings generated by code
                      comment=NA,         ## removes the ## from in front of outputs
                      fig.align="center", ## centers all figures
                      fig.height = 5,     ## set the default height
                      fig.weight = 5      ## set the default width
                      )
```

```{r, include=F}
# libraries
library(arrow)
library(dplyr)
library(ggplot2)
library(pubtheme)
library(patchwork)
```


## Abstract

The report analyzes the swing speed of the batter on fastball pitches
both as an exogenous and endogenous variable. In the exogenous analysis, 
the batter randomly modifies the swing speed as an
independent variable. In the endogenous analysis, the variable 
is treated as an indicator of the batter's strategy given the environment. 

The main implication of the exogenous analysis is that the higher swing speed
is associated with a higher probability of hitting the ball, suggesting that there
There is no optimal swing speed inside the range that maximizes such probability. 
As for the endogenous analysis, the main predictors of the swing speed are 
the score at bat (balls and strikes) and the direction of the pitch (i.e., where the 
ball crosses the strike zone). At the same time, the majority of the swing speed
variance remains unexplained, which gives room for further analysis of the
swing speed as both endogenous and exogenous variables. 



## 1 Introduction 
I cannot emphasize how wrong I was in thinking that baseball is a pretty simple
game. After learning the game's rules this fall, I realized how complex 
and detailed the rules and strategies of the game are. This explains the 
abundance and variety of baseball data and statistical models available. 
At the same time, the data availability is still constantly growing. For example, 
the bat tracking data became publicly available only 
[this year](https://baseballsavant.mlb.com/leaderboard/bat-tracking?_gl=1*4i9lm*_ga*NDYwNDIxODcxLjE3Mjk1Mzc4MjE.*_ga_N8YFCZLYSZ*MTczNDUzNTE1MS4yMi4wLjE3MzQ1MzUxNTEuNjAuMC4xMTE5NjgzMzk1*_gcl_au*MTQyMjg2ODM2MS4xNzI5NTM3ODIx) 
meaning that the analysis of such data is yet to be done. 

This report analyzes the swing speed of batters on fastball pitches. 
The analysis is done in two parts. The swing speed is treated 
as an exogenous variable in the first part. I assume that the batter randomly chooses the swing speed. 
Given that, I analyze how the swing speed affects the probability of hitting the ball into play. 
Another goal in this part is to find out whether there is an optimal swing speed 
within the batter's range, which maximizes the probability of hitting the ball.
This would imply that sometimes the batter is better off swinging the bat
slower to improve the quality of the hit. 

The swing speed is treated as an endogenous variable in the second part of the analysis. 
In this case, the variable can be interpreted as the indicator of the batter's 
cautiousness or confidence. The main goal is to find out the predictors, which 
affect the swing speed the most, and to what extent the environment can influence
the batter's decision to swing the bat faster or slower.

The remainder of the report is structured as follows. In Section 2, I describe 
the dataset I use for the analysis. In Section 3, I describe the data preparation
for the modeling. In Section 4, I present the exogenous analysis, and the 
endogenous one is presented in Section 5. In Section 6, I conclude. Finally, 
I state the directions for further work in Section 7.




## 2 Data 

The data is provided by the 
[CSAS 2025](https://statds.org/events/csas2025/challenge.html) 
Data Challenge. It is pitch-level data for the MLB 2024 season and 
contains 701557 rows from the matches from 4/2/2024 to 10/30/2024. 
Each row represents the pitch and contains 113 variables that describe 
various characteristics of the pitch, including the game, date, inning,
current score in runs, balls and strikes (and outs), players on the field (and bases) on each position, various 
characteristics of the pitch (speed, trajectory, pitch type, spin, etc), and
characteristics of the outcome of the pitch (whether the batter swung, hit, how 
hard he hit, whether he hit the ball into play, the outcome of the pitch, etc). 

```{r}
# reading csv file provided by the organizers
init <- read_csv("statcast_pitch_swing_data_20240402_20241030_with_arm_angle.csv")
```

The main variable in this analysis is `bat_speed` which is the swing speed of the batter. The speed is 
measured at the sweet spot of the bat (around 6 inches from the end of the bat) 
in mph. Note that I have not found the information about at which moment this 
speed is measured (i.e. at the moment of contact with the ball, or at the moment it crosses the strike zone), 
and whether it's some average speed or maximum speed. This information is yet to be found. 
Currently, I assume that the speed is at least measured at the same moment for each pitch. 


## 3 Data Cleaning

First, this report considers the swing speed once the decision to swing was made. 
The analysis of the decision to swing (swing decision score models) is out of the 
scope of this analysis. Thus, I filtered out all the rows where the swing was 
not made. Bunts and 10% of the slowest swings for 
each batter are also filtered out. The reason for doing that is to leave only 'proper swings' in the data. 
By 'proper swings,' I mean the swings where the batter firmly decides to swing
the bat. The bunt is another type of action during the pitch that is more 
suitable for the swing decision score models. The majority of the 
10% of the slowest swings contain very slow swings, where the batter 
changes his mind but already makes enough movement to be counted as a swing. 
About 45% of pitches are swung, so after removing 10% of the slowest
swings, 288630 rows are left in the data. 

Another important decision is to consider only fastballs in this report. 
The swing speed and the outcome of at-bat
heavily depends on the pitch. There are several types of pitches, each of which has unique characteristics and affects the batter's performance
uniquely. Thus, I think that each pitch type should be analyzed 
separately. I decided to analyze the fastballs first. The trajectory of this pitch type
is pretty straight, so these pitches are well 
approximated by the pitch speed, location of the ball, and the horizontal and 
vertical movement of the ball. At the same time, this pitch type is the most
popular. 35% of pitches are fastballs. As for the other pitch types, the more detailed
characteristics of the trajectory of the ball are needed, so one should be 
more careful when dealing with the other pitch types. The analysis of the 
other pitch types can be a part of the further work. 

Finally, I will mainly look at the swing speed normalized by subtracting
the batter's mean from the swing speed. To ensure the demeaning process 
is robust, I sacrifice 2% of the remaining data by filtering out the batters
with less than 30 fastballs properly swung. After that, additional
filtering out of the data that was not correctly measured resulted in 92903 rows 
being left for the modeling. The final cleaned data contains 2354 games and 
509 players who have properly swung at least 30 fastballs in the MLB 2024 season.
This means that the cleaned dataset contains, on average, 182.5 swings per batter.
I use this cleaned dataset in both exogenous and endogenous analyses.



```{r}
## Deprecated cols to delete immediately
init <- init[, !names(init) %in% c("spin_dir", "spin_rate_deprecated", 
                  "break_angle_deprecated", "break_length_deprecated", 
                  "tfs_deprecated", "tfs_zulu_deprecated", "umpire", 
                  "sv_id")]

## Sorting the data, so that: 
# 1. The data is sorted by game_date
# 2. The plays of each game are all together
# 3. In each game the plays are in the order like it was in the actual game
init$game_date <- as.Date(init$game_date, format = "%Y-%m-%d")
init <- init[order(init$game_date, init$home_team, init$away_team, init$at_bat_number, init$pitch_number), ]

# Removing 286 weird rows where lots of missing information
# Looks like the system glitched for these rows
init <- init[!is.na(init$plate_z), ]


## Leaving only the swings (no bunts): 
swings <- init[init$description %in% c("swinging_strike", "swinging_strike_blocked", 
                                     "foul", "hit_into_play", "foul_tip"), ]

## Filtering by events column, leaving only 'proper swings'
swings <- swings[!swings$events %in% c("sac_bunt", "catcher_interf"), ]

## Still some bunts in the data, removing them 
swings <- swings[!(swings$events != "" & grepl("bunt", swings$des, ignore.case = TRUE)), ]


## Converting pitch_type to factor
swings$pitch_type <- as.factor(swings$pitch_type)

## New variable, where I fill NA launch_speed with 0
swings$launch_speed1 <- ifelse(is.na(swings$launch_speed), 0, swings$launch_speed)

## Removing the rows for which the bat_speed and some other
# supporting variables were not correctly measured
# UPD: it's only needed for xWOBA, so don't need this filtering
# if not considering xWOBA, 3% of data
swings <- swings[!is.na(swings$bat_speed), ]
swings <- swings[!((is.na(swings$launch_speed) | is.na(swings$launch_angle)) & 
                     swings$description == 'hit_into_play' & 
                     !(swings$bb_type == 'ground_ball')), ]


## Removing 10% slowest swings for each player to 
# leave only proper swings and remove all bunts which still present in the data
# Analyzing the slow swings where the batter 'changed his mind, but too late'
# FOR FUTURE WORK
swings <- swings |> group_by(batter) |> mutate(q10swing_speed = quantile(bat_speed, 0.1)) |> ungroup()
swings <- swings[swings$bat_speed >= swings$q10swing_speed, ]



## Only fastball swings

# UPD: After some EDA I decided to work with fastballs only
# The swing speed as well as the outcome of at bat
# Heavily depends on the pitch. Every pitch is of different 
# type and lots of characteristics/columns needed to account for the pitch
# Analysis of the swing speed and how it is affected or affects the success 
# of the batter does not make any sense if not account for the pitch. 
# By narrowing the scope to the fastballs only, I can well approximate the pitch
# characteristics by considering just the pitch speed and where the ball landed
# on the strike zone. For the other pitches, I had to use lots of other variables 
# to characterize the curvature of the pitch. So as at this moment I decided 
# to avoid such 'high-dimensionality'. I also expect the results vary greatly 
# between the pitch types, so it makes sense to study data for each pitch 
# type separately. This project will focus on the fastballs only. 
d1 <- swings[swings$pitch_type == "FF", ]


## How many games in total in the data
d1$game_id <- paste(d1$game_date, d1$home_team, d1$away_team)



## Creating summary statistics for each PLAYER/HAND combination 
# There are 55 AMBIDEXTERS in the data, that's why group by stand as well
d1 <- d1 |> 
  group_by(batter, stand) |> 
  mutate(max_bat_speed_player = max(bat_speed), 
         min_bat_speed_player = min(bat_speed), 
         avg_bat_speed_player = mean(bat_speed), 
         sd_bat_speed_player = sd(bat_speed), 
         n_bats = n()
         ) |> 
  ungroup()

## Leaving BATTERS/STAND with at least 30 FF d1 ('solid' properly measured d1)
d1 <- d1[d1$n_bats >= 30, ]

```


```{r}
## Creating aggregate pitchers data on the whole dataset 
# (not filtered based on batters) for the variable creation and analysis

# Normalizing plate_x and plate_z
# plate_x is 0 in the middle of the strike zone
# Let me normalize it so that -1 and 1 are side margins of strike zone
# The width of the strike zone is 17 inches or 17/12 feet
init$plate_x_norm <- (init$plate_x) / (17/12) * 2
# For the plate_z I want to normalize so that 0 is the bottom of the strike zone
# And 5.2 is the top. 5.2 because the height of the strike zone is on average
# 2.6 times bigger than the height
init$plate_z_norm <- (init$plate_z - init$sz_bot) / (init$sz_top - init$sz_bot) * 5.2


# No pitchers ambidexters (only batters ambidexters)
k <- init |> group_by(pitcher) |> 
  summarize(hands = n_distinct(p_throws)) 

# Creating the pitchers data
pitchers_data <- init |> 
  group_by(pitcher) |> 
  summarize(avg_release_speed = mean(release_speed, na.rm = TRUE), 
            avg_release_speed_FF = mean(release_speed[pitch_type == "FF"], na.rm = TRUE), 
            n_pitch = n(), 
            n_pitch_FF = sum(pitch_type == "FF", na.rm = T), 
            p_throws = first(p_throws),
            avg_plate_x_FF = mean(plate_x_norm[pitch_type == "FF"], na.rm = T), 
            avg_plate_z_FF = mean(plate_z_norm[pitch_type == "FF"], na.rm = T),
            avg_plate_x_FF_L = mean(plate_x_norm[pitch_type == "FF" & stand == "L"], na.rm = T),
            avg_plate_z_FF_L = mean(plate_z_norm[pitch_type == "FF" & stand == "L"], na.rm = T),
            avg_plate_x_FF_R = mean(plate_x_norm[pitch_type == "FF" & stand == "R"], na.rm = T),
            avg_plate_z_FF_R = mean(plate_z_norm[pitch_type == "FF" & stand == "R"], na.rm = T),
            avg_pfx_x_FF = mean(pfx_x[pitch_type == "FF"], na.rm = T),
            avg_pfx_z_FF = mean(pfx_z[pitch_type == "FF"], na.rm = T), 
            avg_pfx_x_FF_L = mean(pfx_x[pitch_type == "FF" & stand == "L"], na.rm = T),
            avg_pfx_z_FF_L = mean(pfx_z[pitch_type == "FF" & stand == "L"], na.rm = T),
            avg_pfx_x_FF_R = mean(pfx_x[pitch_type == "FF" & stand == "R"], na.rm = T),
            avg_pfx_z_FF_R = mean(pfx_z[pitch_type == "FF" & stand == "R"], na.rm = T)
            ) |> 
  ungroup() |> 
  arrange(desc(n_pitch_FF))

# 13% of pitchers never pitched fastballs in the season!!!
# nrow(pitchers_data)
# nrow(pitchers_data[pitchers_data$n_pitch_FF == 0, ]) / nrow(pitchers_data)
pitchers_data$prop_pitch_FF <- pitchers_data$n_pitch_FF / pitchers_data$n_pitch


# Merging the pitchers data to d1
d1 <- d1 |> 
  left_join(pitchers_data, by = c("pitcher" = "pitcher", "p_throws" = "p_throws"))


```



## 4 Swing Speed as an Exogenous Variable

In this section, I analyze how the swing speed affects the batter's performance. 
The probability of hit into play (i.e., hitting the ball into the fair zone) is used as
the metric for such quality. I assume that the swing speed is exogenous, i.e. 
the batter randomly chooses the swing speed. This assumption is needed to 
make some causal statements and suggest the optimal swing speed. This assumption
is challenged in the next section, where the swing speed is treated as an
endogenous variable. 

In Section 4.1, I present the models and the results. Before the modeling,
I did some exploratory data analysis which helped me to create the variables 
used in the models. You can find the main results (visualizations and comments)
from this exploratory data analysis in Appendix A. 




```{r}
### Variables creation for the models in section 4.1
# Please check Appendix A for more explanations, visualizations and comments

d1$xWOBA1 <- ifelse(is.na(d1$estimated_woba_using_speedangle), 0, 
                   d1$estimated_woba_using_speedangle)
d1$TARGET <- ifelse(d1$description == "hit_into_play", 1, 0)
d1$TARGET_f <- as.factor(d1$TARGET)
d1$bat_speed_norm <- d1$bat_speed - d1$avg_bat_speed_player
d1$bat_speed_sq <- d1$bat_speed^2
d1$bat_speed_norm_sq <- d1$bat_speed_norm^2
d1$plate_x_norm <- (d1$plate_x) / (17/12) * 2
d1$plate_z_norm <- (d1$plate_z - d1$sz_bot) / (d1$sz_top - d1$sz_bot) * 5.2
d1$plate_x_norm_inv <- ifelse(d1$stand == "L", -d1$plate_x_norm, d1$plate_x_norm)
d1$plate_x_norm_inv_less_minus1 <- d1$plate_x_norm_inv * (d1$plate_x_norm_inv < -1)
d1$plate_z_norm_less_0 <- d1$plate_z_norm * (d1$plate_z_norm < 0)
d1$plate_z_norm_norm <- d1$plate_z_norm - mean(d1$plate_z_norm)
d1$pfx_x_inv <- ifelse(d1$stand == "L", -d1$pfx_x, d1$pfx_x)
d1$diff_hand <- ifelse(d1$p_throws != d1$stand, 1, 0)
d1$pfx_x_inv_norm <- d1$pfx_x_inv - mean(d1$pfx_x_inv[d1$diff_hand == 0])
d1$pfx_z_norm <- d1$pfx_z - mean(d1$pfx_z)
d1$strikes2 <- ifelse(d1$strikes == 2, 1, 0)
```



### 4.1 Modelling and Results

#### 4.1.1 Simple aggregate model. 

```{r}
# function for a nicer visualization of the regression results
regr_nice <- function(model){
  st <- summary(model)$coefficients
  var_names <- names(st[, 1])
  
  coefs <- sub("^$", "0", sub("0+$", "", signif(st[, 1], 2)))
  SEs <- sub("^$", "0", sub("0+$", "", signif(st[, 2], 2)))
  pvals <- sub("^$", "0", sub("0+$", "", signif(st[, 4], 2)))
  reg_df <- data.frame(Variable = var_names, Coefficient = coefs, SE = SEs, 
                      pvalue = pvals)
  
  rownames(reg_df) <- 1:nrow(reg_df)
  return(reg_df)
}
```


First, let me introduce the simple model, which was run on all the data simultaneously. The only reason to run this simple model (with potential bias) is its easy interpretability. 


```{r}
## Modelling: simple aggregate model

# Let's create new variable for the better interpretation of the coefficients of the model
d1$release_speed_norm <- d1$release_speed - mean(d1$release_speed)

model1 <- glm(TARGET ~ bat_speed_norm * release_speed_norm + bat_speed_norm_sq, data = d1, family = binomial)
print(model1$call)
regr_nice(model1)
```
The coefficients of this model are: 

- bat_speed_norm: swing speed normalized by subtracting the batter's average swing speed (in mph)
- release_speed_norm: pitch speed normalized by subtracting aggregate average pitch speed in the dataset (in mph)
- bat_speed_norm_sq: square of the normalized swing speed
- bat_speed_norm:release_speed_norm: interaction term between normalized swing speed and normalized pitch speed

Interpretation of the coefficients: 
Note that this is the logit model for the binary outcome, not a linear regression. Thus, the coefficients are not directly interpretable. Still, the signs of the coefficients can be interpreted. If the ball is pitched at the average speed (IF WE FIX THE PLAYER), the batter's increase in swing speed is associated with the increase in the probability of hitting the ball into play. The effect gets slightly smaller if the pitch is thrown at a higher speed, but the effect never turns positive in the reasonable range of pitch speed (<105mph). The effect of the pitch speed is negative no matter what the swing speed is relative to the batter's average. The coefficient to the squared normalized swing speed is insignificant, so there is no evidence of the existence of the optimal swing speed. 

Below, you can see the association between normalized swing speed and `TARGET` for different values of pitch speed.


```{r out.width="50%"}
## Using same model but unnormalized release speed for better visualization
# of the model predictions below

model3 <- glm(TARGET ~ bat_speed_norm * release_speed + bat_speed_norm_sq, data = d1, family = binomial)

preds_d90 <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000), 
                      release_speed = 90)
preds_d92_5 <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000),
                      release_speed = 92.5)
preds_d95 <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000),
                      release_speed = 95)
preds_d97_5 <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000),
                      release_speed = 97.5)
preds_d100 <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000),
                      release_speed = 100)
preds_d90$bat_speed_norm_sq <- preds_d90$bat_speed_norm^2
preds_d92_5$bat_speed_norm_sq <- preds_d92_5$bat_speed_norm^2
preds_d95$bat_speed_norm_sq <- preds_d95$bat_speed_norm^2
preds_d97_5$bat_speed_norm_sq <- preds_d97_5$bat_speed_norm^2
preds_d100$bat_speed_norm_sq<- preds_d100$bat_speed_norm^2

preds_d90$pred <- predict(model3, newdata = preds_d90, type = 'response')
preds_d92_5$pred <- predict(model3, newdata = preds_d92_5, type = 'response')
preds_d95$pred <- predict(model3, newdata = preds_d95, type = 'response')
preds_d97_5$pred <- predict(model3, newdata = preds_d97_5, type = 'response')
preds_d100$pred <- predict(model3, newdata = preds_d100, type = 'response')


g0 <- ggplot(d1, aes(x = bat_speed_norm, y = TARGET)) +
  geom_line(data = preds_d90, aes(x = bat_speed_norm, y = pred, color = '90'), size = 1) +
  geom_line(data = preds_d92_5, aes(x = bat_speed_norm, y = pred, color = '92.5'), size =1) +
  geom_line(data = preds_d95, aes(x = bat_speed_norm, y = pred, color='95'), size = 1) +
  geom_line(data = preds_d97_5, aes(x = bat_speed_norm, y = pred, color='97.5'), size = 1) +
  geom_line(data = preds_d100, aes(x = bat_speed_norm, y = pred, color='100'), size = 1) +
  scale_color_manual(name = "Pitch Speed", values = c("90" = "green", 
            "92.5"="blue", "95"='red', "97.5"='purple', "100"='yellow'), 
            breaks = c("90", "92.5", "95", "97.5", "100")) + 
  labs(title = "Predicted Target vs Normalized Swing Speed \n Given Pitch Speed", 
      x = "Normalized (demeaned) Swing Speed, mph", y = "Target") + 
  theme_grey(base_size = 16)

g0
```



#### 4.1.2 Simple models for each player separately.

```{r}
## Getting data with players' characteristics

# Saving info about ambidexters and dominant stand
d1 <- d1 |> group_by(batter, player_name) |> 
  mutate(ambidexter = ifelse(sum(stand == "L") > 0 & sum(stand == "R") > 0, 1, 0), 
         dominant_stand = ifelse(sum(stand == "L") > sum(stand == "R"), "L", "R")) |> 
  ungroup()

# Dataset with the players' information
df_players1 <- d1 |> group_by(batter, player_name, dominant_stand) |>
  summarise(min_speed = min(bat_speed), max_speed = max(bat_speed),
            avg_speed = mean(bat_speed), sd_speed = sd(bat_speed),
            n_swings = n(), avg_TARGET = mean(TARGET),
            avg_pitch_speed = mean(release_speed), avg_xwoba = mean(xWOBA1), 
            ambidexter = mean(ambidexter)) |>
  ungroup() |>
  arrange(desc(avg_speed))

```

```{r}
## Fitting the model above FOR EACH BATTER INDIVIDUALLY

options(dplyr.summarise.inform = FALSE)

df_coef_p <- data.frame()
colnames_df_coef_p <- c("batter", "player_name", "dominant_stand", "min_speed", 
                         "max_speed", "avg_speed", "sd_speed", "n_swings", 
                         "avg_TARGET", "avg_pitch_speed", "avg_xwoba", "ambidexter", 
                         "intercept", "coef_bs", "coef_ps", "coef_bs_sq", "coef_bs_ps",
                         "p_bs", "p_ps", "p_bs_sq", "p_bs_ps")

# getting needed info player by player
k <- nrow(df_players1)
for (i in 1:k){
  PLAYER_INFO <- df_players1[i, ]
  d1p <- d1[d1$batter == PLAYER_INFO$batter, ]
  
  # Using the release speed normalized by global average
  # But the swing speed is normalized by the player's average
  d1p$bat_speed_norm_p <- d1p$bat_speed - mean(d1p$bat_speed)
  d1p$bat_speed_norm_p_sq <- d1p$bat_speed_norm_p^2
  
  model1 <- glm(TARGET ~ bat_speed_norm_p * release_speed_norm + bat_speed_norm_p_sq, data = d1p, family = binomial)
  
  COEF_VEC <- c(summary(model1)$coefficients[, 1], summary(model1)$coefficients[2:5, 4])
  
  # Final row of the dataframe
  OPT_FULL <- c(PLAYER_INFO, COEF_VEC)
  names(OPT_FULL) <- paste("X", 1:length(OPT_FULL), sep = '')
  df_coef_p <- rbind(df_coef_p, OPT_FULL)
}
colnames(df_coef_p) <- colnames_df_coef_p
```


```{r}
# I think 100 swings is enough for the model with 3 vars (and intercept)
# 371 players (out of 515) have >100 swings (solid, properly recorded, FF pitch)
# I'll look at the players with >100 swings
# Also, look at the left-hand and right-hand batters separately: 
# UPD: in final report I consider them together, you can mention that 
# lh and rh results are same


df_coef_p1 <- df_coef_p |> filter(n_swings > 100)
```


I ran the model above on each player separately. I considered only 370 out of 509 players with more than 100 swings. There is not enough data for the other players to fit the model with 5 variables. It turns out that almost for any player, this simple model has too many variables to make any conclusions. The only significant result is a pretty stable positive coefficient to the normalized swing speed, suggesting that normalized swing speed is not just on average associated with the increase in `TARGET,` but this is the case for almost every player. The other coefficients are noisy and not significant, and no clusterization based on the coefficients' values was observed. Also, the average swing speed of the player had no correlation with any of the coefficients of the model, which might imply that the simple model run above generalizes the effects well for each player, but little or no evidence can be provided due to a small number of observations per player. In the future, an additional attempt can be made by carefully adding other pitch types. 

```{r fig.width = 12, out.width="80%"}
## Distribution of the coefficients across batters
# Not significant, but slightly positive on average. 
# At the average pitch speed, bat speed increase increases the probability of contact

k1 <- ggplot(df_coef_p1, aes(x = coef_bs)) + 
  geom_histogram(bins = 30, fill = 'blue') + 
  labs(title = 'bat_speed_norm Coefficients Distribution \n (per batter)', 
       x = 'Coefficient Value') + 
  theme_grey(base_size = 16)
k2 <- ggplot(df_coef_p1, aes(x = p_bs)) + 
  geom_histogram(bins = 30, fill = 'blue') + 
  labs(title = ' p-values Distribution, Coefficients to \n bat_speed_norm (per batter)', 
       x = 'p-value') + 
  theme_grey(base_size = 16)

k1 + k2

```


#### 4.1.3 Final aggregate model.

If all of the variables from Section 4.2.1 are considered, the following model is proposed.

```{r}
## Using the variables created above fitting the final model in the exogenous
# analysis
model1 <- glm(TARGET ~ bat_speed_norm + bat_speed_norm_sq +  release_speed_norm + 
                plate_x_norm_inv + plate_x_norm_inv_less_minus1 + 
                plate_z_norm_norm + plate_z_norm_less_0 + 
                diff_hand + pfx_x_inv_norm + pfx_z_norm + 
                balls + strikes2 + 
                bat_speed_norm:(release_speed_norm + 
                plate_x_norm_inv + plate_x_norm_inv_less_minus1 + 
                plate_z_norm_norm + plate_z_norm_less_0 + 
                diff_hand + strikes2), data = d1, family = binomial)
print(model1$call)
regr_nice(model1)
```


The coefficients of this model are (some of the variables are a little weird so that I can better interpret the model later): 

- bat_speed_norm: swing speed normalized by subtracting the batter's average swing speed (in mph)
- bat_speed_norm_sq: square of the normalized swing speed
- release_speed_norm: pitch speed normalized by subtracting aggregate average pitch speed in the dataset (in mph)
- plate_x_norm_inv: inverted (multiplied by -1 if left stand) normalized horizontal location of the ball 
- plate_x_norm_inv_less_minus1: interaction term between the inverted normalized horizontal location of the ball and the indicator of the inverted normalized horizontal location is less than -1
- plate_z_norm_norm: normalized vertical location of the ball normalized (one more time) by subtracting the average vertical location of the ball
- plate_z_norm_less_0: interaction term between the normalized vertical location of the ball and the indicator of the normalized vertical location is less than 0
- diff_hand: indicator of the different hand of the pitcher and stand of the batter
- pfx_x_inv_norm: inverted horizontal movement of the ball normalized by subtracting the average inverted horizontal movement of the ball
- pfx_z_norm: vertical movement of the ball normalized by subtracting the average vertical movement of the ball
- balls: number of balls in the at-bat
- strikes2: an indicator of the number of strikes in the at-bat is 2
- bat_speed_norm:release_speed_norm: interaction term between bat_speed_norm and release_speed_norm
- bat_speed_norm:plate_x_norm_inv: interaction term between normalized swing speed and plate_x_norm_inv
- bat_speed_norm:plate_x_norm_inv_less_minus1: interaction term between normalized swing speed and plate_x_norm_inv_less_minus1
- bat_speed_norm:plate_z_norm_norm: interaction term between bat_speed_norm speed and plate_z_norm_norm
- bat_speed_norm:plate_z_norm_less_0: interaction term between bat_speed_norm speed and plate_z_norm_less_0
- bat_speed_norm:diff_hand: interaction term between bat_speed_norm speed and diff_hand
- bat_speed_norm:strikes2: interaction term between bat_speed_norm speed and strikes2

In the model interpretation, I will only focus on the `bat_speed_norm.` If the ball is pitched at the average speed, in the average location, the pitcher and the batter have the same hand, the movement of the ball is average, the number of balls is 0, and the number of strikes is 2, the effect of the normalized swing speed on the probability of the 
hit into play is positive, and is very similar to the one in the simple model, so you can refer to the graphs of the fitted probabilities in the simple model to visualize this effect. Moreover, despite the significant interaction terms, the effect of the normalized swing speeds stays positive. For example, let us consider one of the 'worst-case scenarios.' If the ball is pitched at 100mph and thrown at the opposite side edge of the strike zone from the pitcher with the opposite hand, the score at bat is 0 balls and 2 strikes, then the coefficient to the `bat_speed_norm` will drop from 0.086 to 0.015, but still will be positive.  

 
Note that you can find my comment about using xwOBA as a target (instead of the probability of the hit) in Appendix B.


## 5 Swing Speed as an Endogenous Variable

In this section, I present the swing speed analysis as an Endogenous Variable. 
I assume that the environment (the stage of the game) makes the batter adapt his strategy by 
changing the swing speed (or simply some aspects of the game force him to change the swing speed). 
I use only aggregate analysis in this section. Since it makes more sense to consider how the swing speed is changed relative to the 
batter's average, I use the normalized swing speed from the previous section.

Like in the previous section, I present the models and the results in Section 5.1. 
The results (visualizations and comments) of the exploratory data analysis which
helped me to create the variables used in the models can be found in Appendix C. 



```{r}
### Variables creation for the models in section 5.1
# Please check Appendix C for more explanations, visualizations and comments

d1$release_speed_norm <- d1$release_speed - d1$avg_release_speed_FF
d1$inning1 <- ifelse(d1$inning <= 9, d1$inning, '>9')
d1$inning2 <- ifelse(d1$inning <= 9, d1$inning, 10)
d1$inning3 <- ifelse(d1$inning1 %in% c("2", "3", "4", "5"), "2-5", d1$inning1)
d1$top_inning7plus <- ifelse(d1$inning2 >= 7 & d1$inning_topbot == "Top", 1, 0)
d1$b1_loaded <- ifelse(!is.na(d1$on_1b), 1, 0)
d1$b2_loaded <- ifelse(!is.na(d1$on_2b), 1, 0)
d1$b3_loaded <- ifelse(!is.na(d1$on_3b), 1, 0)
d1$bases_loaded <- paste0(d1$b3_loaded, d1$b2_loaded, d1$b1_loaded)
d1$bases_loaded1 <- d1$b1_loaded + d1$b2_loaded + d1$b3_loaded
d1$score_at_bat <- paste0(d1$balls, "/", d1$strikes)
d1$plate_x_norm_inv <- ifelse(d1$stand == "L", -d1$plate_x_norm, d1$plate_x_norm)
d1$plate_x_norm_inv_less0 <- d1$plate_x_norm_inv * (d1$plate_x_norm_inv < 0)
d1$plate_z_norm_less1_5 <- d1$plate_z_norm * (d1$plate_z_norm < 1.5)
d1 <- d1 |> group_by(batter, stand) |>
  mutate(mean_sz_bot = mean(sz_bot),
         mean_sz_top = mean(sz_top))
d1$sz_top_demean <- d1$sz_top - d1$mean_sz_top
d1$sz_bot_demean <- d1$sz_bot - d1$mean_sz_bot
```



### 5.2 Modelling and Results

Before running the final model, I want to mention that the number of outs in the inning and the score in runs are surprisingly insignificant (I tried lots of variables' variations and interactions with the other covariates). Also, the interaction of the final variables in the models did not improve the fit, nor did the use of the lightgbm model instead of the linear regression model. 

```{r}
## Forming the final model and fitting

# variable for better interpretability and fit
d1$release_speed_minus_avg_release_speed_FF <- d1$release_speed - d1$avg_release_speed_FF
d1$inning_final <- ifelse(d1$inning3 %in% c('6', '7', '8', '9', '>9'), '6+', d1$inning3)
d1$inning_final <- relevel(as.factor(d1$inning_final), ref = "1")

# Outs when up, score differential: not important
# score differential zero is significant, but correlated with first inning
# Could use fancier variable, which gives slight increase in R^2
# I wanted to deal with avg_plate_x_FF and avg_plate_z_FF, but
# That's for future work, significant, but not impactful on averall fit
# so not a big deal (real small effect) + very tricky to deal with and needed
# to create lots of more variables
model_final <- lm(bat_speed_norm ~ release_speed_minus_avg_release_speed_FF + avg_release_speed_FF + 
             inning_final + 
             top_inning7plus + balls + strikes + 
             b1_loaded + b2_loaded + b3_loaded +
             plate_x_norm_inv + plate_x_norm_inv_less0 + 
             plate_z_norm + plate_z_norm_less1_5 + 
             pfx_x:stand:p_throws + pfx_z:stand + 
             sz_bot_demean, data = d1)
print(model_final$call)
regr_nice(model_final)
paste0("R^2: ", round(summary(model_final)$r.squared, 3))

# Score_at_bat
# summary(lm(bat_speed_norm ~ release_speed_minus_avg_release_speed_FF + avg_release_speed_FF + 
#              inning_final + 
#              top_inning7plus + balls + as.factor(score_at_bat) + 
#              b1_loaded + b2_loaded + b3_loaded +
#              plate_x_norm_inv + plate_x_norm_inv_less0 + 
#              plate_z_norm + plate_z_norm_less1_5 + 
#              pfx_x:stand:p_throws + pfx_z:stand + 
#              sz_bot_demean, data = d1))


```

Variables of the model are as follows:

- release_speed_minus_avg_release_speed_FF: pitch speed normalized by subtracting the average fastball pitch speed of the pitcher (in mph)
- avg_release_speed_FF: average fastball pitch speed of the pitcher (in mph)
- inning_final: factor variable of the inning. Three levels: "1" (reference category), "2-5", "6+"
- top_inning7plus: indicator of the top half-inning and the inning is 7 or later
- balls: number of balls in the at-bat
- strikes: number of strikes in the at-bat
- b1_loaded: indicator of the first base loaded
- b2_loaded: indicator of the second base loaded
- b3_loaded: indicator of the third base loaded
- plate_x_norm_inv: inverted (multiplied by -1 if left stand) normalized horizontal location of the ball (from the catcher's perspective)
- plate_x_norm_inv_less0: interaction term between the inverted normalized horizontal location of the ball and the indicator of the inverted normalized horizontal location is less than 0
- plate_z_norm: normalized vertical location of the ball
- plate_z_norm_less1_5: interaction term between the normalized vertical location of the ball and the indicator of the normalized vertical location is less than 1.5
- sz_bot_demean: vertical location of the bottom of the strike zone normalized by subtracting the average vertical location of the bottom of the strike zone of the batter
- standL:pfx_z: interaction term between dummy for the left stand and the vertical movement of the ball (from the catcher's perspective)
- standR:pfx_z: interaction term between the dummy for the right stand  and the vertical movement of the ball (from the catcher's perspective)
- pfx_x:standL:p_throwsL: horizontal movement of the ball (from the catcher's perspective) * dummy for the left stand * dummy for the left-handed pitcher
- pfx_x:standL:p_throwsR: horizontal movement of the ball (from the catcher's perspective) * dummy for the left stand * dummy for the right-handed pitcher
- pfx_x:standR:p_throwsL: horizontal movement of the ball (from the catcher's perspective) * dummy for the right stand * dummy for the left-handed pitcher
- pfx_x:standR:p_throwsR: horizontal movement of the ball (from the catcher's perspective) * dummy for the right stand * dummy for the right-handed pitcher


Interpretation:
Almost all of the coefficients are of the expected sign. Pitch speed negatively affects the normalized swing speed. The later the innings, the lower the normalized swing speed is on average. Away teams start to swing slower than the home teams after inning 6. The number of balls has a positive association, and the number of strikes has a negative one. Each base loaded decreases the normalized swing speed. The first base has the smallest effect; the third base has the highest one. Keeping everything else constant, the normalized swing speed is highest when the ball is in the middle of the plate (horizontal-wise) and as low as possible vertical-wise. The lower squat of the batter is slightly associated with the decrease in the normalized swing speed. More vertical movement downwards is associated with a lower normalized swing speed. For a right-hand batter, if the pitcher is left-handed, the normalized swing speed is higher if the ball moves away from him horizontally. If the pitcher is right-handed, the effect is the opposite. The left-hand batter swings faster if the ball moves away from him horizontally (this one is unexpected). 



## 6 Conclusion

In conclusion, this report shows the swing speed analysis as both exogenous and endogenous variable. The analysis of the swing speed as an exogenous variable shows that the probability of hit into play is, on average, positively related to the swing speed once we fix the batter. The effect stays positive at almost all feasible values of the other control variables. The coefficient to the squared term is significant in neither of the models presented, suggesting the absence of the optimal swing speed. It can be suggested that the increased swing speed causes an increase in the probability of the hit into play for the batter. However, it can be the case of confounding variables, such as the 'confidence' of the batter, which can be affected by the 'easiness' of the pitch. In other words, it is not the case that the batter randomly chooses the swing speed, and increased swing speed causes a better quality of the hit. It is more likely that the batter chooses a higher swing speed because he thinks that he is more prepared for this pitch and feels more confident to swing it at a higher speed. Since he is more prepared, he is more likely to hit the ball into play. Thus, there might be no causal relationship between the swing speed and the probability of the hit into play.

As for the swing speed analysis as an endogenous variable, the batter explains 36% of the variability in swing speed. If we normalize the swing speed, the main predictors of the normalized swing speed are the number of balls, number of strikes, and location of the ball. Note that the model explains only 18.5% of the variability of the normalized swing speed. This means that either there are some other factors missed in the analysis, the noise is relatively high (might be measurement errors), and/or the swing speed is mostly exogenous and is mostly determined by the batter at his discretion, independent of the environment. I think that the first two points are the most likely, and more analysis can be done in future work. 

## 7 Future Work

The analysis reported can be extended further. In fact, this analysis seeds more questions than answers. A more detailed causal inference analysis can be done for the exogenous part of the analysis. At the same time, a more robust analysis can be created to model the quality of the hit for each batter individually. Furthermore, other measures of the quality of the hit can be considered (i.e., xWOBA). I have also done probability of contact (any contact with the ball). The results were pretty much the same as for the probability of getting hit into play. Analyzing all pitch types (not just fastballs) might help with robustness in both parts of the analysis. As for the endogenous part of the analysis, more variables can be derived to improve the model (i.e., the autocorrelation of the swing speed per batter, a measure of the tiredness of the batter, and an analysis of the extreme values of the normalized swing speed). At the same time, more research can be done to look for any current analysis on the swing speed (since it is new data, there might be a new analysis/information every day). More qualitative analysis can be done. For example, I would like to find and ask people related to baseball what makes them swing at different speeds. Also, I could include broader aspects of the bat tracking data (i.e., bat length, which is also newly available data). Finally, it could be useful to consider not only proper swing data and merge the current analysis with the swing decision model since the factors affecting the decision to swing (which I might miss in this analysis) might similarly affect the swing speed.   



## Appendix 

### Appendix A: Exploratory Data Analysis, exogenous analyis.


```{r}
# adding the initial target variable: expected WOBA
d1$xWOBA1 <- ifelse(is.na(d1$estimated_woba_using_speedangle), 0, 
                   d1$estimated_woba_using_speedangle)

```




```{r}
# Creating target variable, 1 if hit into play, 0 otherwise
# Only 34% of the swings result in the hit into play
d1$TARGET <- ifelse(d1$description == "hit_into_play", 1, 0)
d1$TARGET_f <- as.factor(d1$TARGET)
```

To begin with, my target variable for the exogenous analysis is the binary
variable, which equals one if the batter hits the ball into the play and zero
otherwise. The variable is called `TARGET`. In the cleaned data, 34%
of the swings resulted in the hit into play. 

In this subsection, I will present the variables which are correlated 
with the `TARGET` variable, which will be used in the modelling. 

Let me start with the swing speed. In the first histogram below, you can see 
the variability in the average swing speed of the batters. To be able to 
conduct a more interpretable aggregate analysis (with all of the players at the same time), 
I created the normalized swing speed variable.  


```{r}
# First let's create some summary stat for each player + demeaned swing speed
d1$bat_speed_norm <- d1$bat_speed - d1$avg_bat_speed_player
```


```{r fig.width=12, out.width="80%"}
# Distribution of swing speed

d1p <- d1 |> group_by(batter) |>
  summarize(avg_TARGET = mean(TARGET), 
            avg_bat_speed = mean(bat_speed), 
            n = n()) |> 
  ungroup()

k1 <- ggplot(d1p, aes(x = avg_bat_speed)) +
  geom_histogram(bins = 20, fill = 'blue') +
  labs(title = "Average Swing Speed Distribution (per batter)", x = "Average Swing Speed per batter, mph") + 
  theme_grey(base_size = 16)
k2 <- ggplot(d1p, aes(x = n)) +
  geom_histogram(bins = 20, fill = 'blue') +
  labs(title = "Swing Number Distribution (per batter)", x = "Number of Swings, per batter")  + 
  theme_grey(base_size = 16)
k3 <- ggplot(d1, aes(x = bat_speed)) +
  geom_histogram(bins = 50, fill = 'blue') +
  labs(title = "Swing Speed Distribution", x = "Swing Speed, mph") + 
  theme_grey(base_size = 16)
k4 <- ggplot(d1, aes(x = bat_speed_norm)) +
  geom_histogram(bins = 50, fill = 'blue') +
  labs(title = "Normalized Swing Speed Distribution", x = "Swing Speed, mph") + 
  theme_grey(base_size = 16)

(k1 + k2) 

(k3 + k4)



# average sd per player
k <- d1 |> group_by(batter) |> 
  summarize(avg_sd = sd(bat_speed))
```

```{r}
## Logit models

# raw swing speed
model1 <- glm(TARGET ~ bat_speed, data = d1, family = binomial)
# summary(model1)
preds <- data.frame(bat_speed = seq(min(d1$bat_speed), max(d1$bat_speed), 
                                    length.out = 1000))
preds$pred <- predict(model1, newdata = preds, type = 'response')

# demeaned swing speed
model2 <- glm(TARGET ~ bat_speed_norm, data = d1, family = binomial)
# summary(model2)
preds_z <- data.frame(bat_speed_norm = seq(min(d1$bat_speed_norm), 
                                             max(d1$bat_speed_norm), 
                                             length.out = 1000))
preds_z$pred <- predict(model2, newdata = preds_z, type = 'response')
```

The player with the higher average swing speed tends to have less proportion of balls hit 
into play. However, the relationship between `TARGET` and normalized swing speed is positive. 
This means that if we fix some batter, then on average, the higher swing speed is associated with 
the higher probability of hitting the ball into play. Moreover, the latter effect dominates so that, 
in general, the swing speed is associated with the higher probability of hitting the ball into play even 
if we do not fix some batter. 

At the same time, no apparent concavity is noticed at this point, so it might not be the 
case that the optimal swing speed exists. 

```{r fig.width=15, out.width="110%", out.align='left'}
#Target vs raw swing speed scatter

# add titles and axis labels
g1 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = bat_speed, y = TARGET)) +
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_line(data = preds, aes(x = bat_speed, y = pred), color = 'blue') +
  labs(title = "Target vs Swing Speed", x = "Swing Speed, mph", y = "Target") +
  coord_cartesian(ylim = c(-.1, 1.1)) + 
  theme_grey(base_size = 16)


# Target vs swing speed demeaned scatter
g2 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = bat_speed_norm, y = TARGET)) +
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  labs(title = "Target vs Normalized Swing Speed", x = "Normalized Swing Speed, mph", y = "Target") +
  geom_line(data = preds_z, aes(x = bat_speed_norm, y = pred), color = 'blue')  + 
  theme_grey(base_size = 16)



# Avg target per player vs avg swing speed per player

g3 <- ggplot(d1p, aes(x = avg_bat_speed, y = avg_TARGET)) +
  geom_point(alpha = 1, size = 1) +
  geom_smooth(color = 'red') +
  labs(title = "Avg Target vs Avg Swing Speed \n (per batter)", x = "Batter's average swing speed, mph", y = "Batter's average Target") +
  geom_smooth(method = 'lm', color = 'blue') + 
  theme_grey(base_size = 16)+
  theme(plot.title = element_text(size = 16))

g1 + g2 + g3
```


```{r}
# Creating squared term for optimality
d1$bat_speed_sq <- d1$bat_speed^2
d1$bat_speed_norm_sq <- d1$bat_speed_norm^2
```


The fastball's release speed (the pitch speed) is more stable between the pitchers than the 
swing speed between the batters. As expected, the release speed is negatively associated with the 
probability of hitting the ball into play. 

```{r fig.width=8, fig.height=4, out.width="80%"}

## Release (pitch) speed

# hist(d1$release_speed, breaks = 100, col = 'blue', 
#     main = 'Pitch Speed Distribution (fastballs)', 
#     xlab = 'Pitch speed, mph')

# redo above histogram using ggplot
g1 <- ggplot(d1, aes(x = release_speed)) +
  geom_histogram(bins = 50, fill = 'blue') +
  labs(title = "Pitch Speed Distribution", x = "Pitch Speed, mph") + 
  theme_grey(base_size = 16)

model1 <- glm(TARGET ~ release_speed, data = d1, family = binomial)
# summary(model1)
preds <- data.frame(release_speed = seq(min(d1$release_speed), max(d1$release_speed), 
                                    length.out = 1000))
preds$pred <- predict(model1, newdata = preds, type = 'response')

# linear (logit) model seems like a good fit
g0 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = release_speed, y = TARGET)) +
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_line(data = preds, aes(x = release_speed, y = pred), color = 'blue') +
  labs(title = "Target vs Pitch Speed", x = "Pitch Speed, mph", y = "Target") + 
  theme_grey(base_size = 16)

g1 + g0 
```



The location of the ball (the point where it crosses the strike zone) 
heavily depends on the pitcher's throwing hand and the batter's stand. 
The diagrams below represent the strike zone from the catcher's 
perspective. Thus, it can be seen that, on average, both left-hand and 
right-hand batters receive the ball on the opposite side of the strike zone. 
At the same time, right-hand and left-hand pitchers throw the balls to the 
different regions of the strike zone when pitching to right-hand batters.
Thus, it is important to look at each of these four cases separately (l to r, 
r to r, l to l, r to l) 
before determining the association between the location of the ball and 
the probability of hitting the ball into play. 

Note that in the diagrams below and further, I refer to the ball's 'Normalized Vertical/Horizontal' location. The normalized version of the location of the 
ball is derived such that the normalized horizontal location of the left edge of the strike zone (from the catcher's 
perspective) is equal to -1, and the right edge is equal to 1, and the 
the normalized vertical location of the bottom edge of the strike zone is equal to 0, and the
top edge is equal to 5.2 (this weird 5.2 value is because the height of the strike
zone is, on average, 2.6 higher than the width of the strike zone). 


```{r, fig.width=12, fig.height=6, out.width="80%"}
## Vertical and horizontal location of the pitch

# Analyzing where pitchers throw the FF pitch on average (pitchers with >50 FF pitches)
pitchers_data1 <- pitchers_data |> filter(n_pitch_FF > 50)


g1 <- ggplot(pitchers_data1, aes(x = avg_plate_x_FF_L, y = avg_plate_z_FF_L, color = p_throws)) +
  geom_point(alpha = 0.7) + 
  geom_segment(aes(x = -1, y = 0, xend = -1, yend = 5.2), color = 'black', size = 1) + 
  geom_segment(aes(x = 1, y = 0, xend = 1, yend = 5.2), color = 'black', size = 1) + 
  geom_segment(aes(x = -1, y = 0, xend = 1, yend = 0), color = 'black', size = 1) + 
  geom_segment(aes(x = -1, y = 5.2, xend = 1, yend = 5.2), color = 'black', size = 1) + 
  scale_colour_manual(values = c("red","blue")) + 
  labs(title = "Average Pitch Location to \n Left-Hand Batters (per pitcher)", 
       x = "Average Horizontal Ball Location (per pitcher)", y = "Average Vertical Ball Location (per pitcher)") +
  guides(color=guide_legend(title="Pitch Hand")) + 
   coord_fixed(ratio = 0.5, ylim = c(0, 6), xlim = c(-1.2, 1.2)) + 
  theme_grey(base_size = 14)

g2 <- ggplot(pitchers_data1, aes(x = avg_plate_x_FF_R, y = avg_plate_z_FF_R, color = p_throws)) +
  geom_point(alpha = 0.7) + 
  geom_segment(aes(x = -1, y = 0, xend = -1, yend = 5.2), color = 'black', size = 1) + 
  geom_segment(aes(x = 1, y = 0, xend = 1, yend = 5.2), color = 'black', size = 1) + 
  geom_segment(aes(x = -1, y = 0, xend = 1, yend = 0), color = 'black', size = 1) + 
  geom_segment(aes(x = -1, y = 5.2, xend = 1, yend = 5.2), color = 'black', size = 1) + 
  scale_colour_manual(values = c("red","blue")) + 
  labs(title = "Average Pitch Location to \n Right-Hand Batters (per pitcher)", 
       x = "Average Normalized Horizontal Ball Location (per pitcher)", y = "Average Normalized Vertical Ball Location (per pitcher)") +
  guides(color=guide_legend(title="Pitch Hand")) + 
   coord_fixed(ratio = 0.5, ylim = c(0, 6), xlim = c(-1.2, 1.2)) + 
  theme_grey(base_size = 14)

g1 + g2

# pitchers_data1 |> group_by(p_throws) |> 
#   summarize(avg_plate_x_FF_L = mean(avg_plate_x_FF_L), 
#             avg_plate_x_FF_R = mean(avg_plate_x_FF_R),
#             avg_plate_z_FF_L = mean(avg_plate_z_FF_L), 
#             avg_plate_z_FF_R = mean(avg_plate_z_FF_R))



# Plate_z, plate_x: pitchers pitch away from the batters, i.e. more to the left 
# (from the catchers perspective) to lefties and to the right to righties. 
# With the exception that left-hand pitchers pitch to the middle of strike zone
# to righties (I don't know why). Also, pitchers pitch to the opposite hand 
# higher than to the same hand, but there's still some difference between 
# rh and lh pitchers. 

# Thus, I propose to look at 4 scenarios separately: 
# l to r, r to l, l to l, r to r and create the variables accordingly

```


After looking at each of the 4 cases, I conclude that the hand of the pitcher
does not affect the effect of the ball's location. However, the stand
affects how the horizontal location of the ball affects the probability of hit 
into play. We can see that if the ball is thrown in the strike zone, the further
it is from the batter, the less likely the batter hits the ball into play.
The effect is the opposite once the ball is thrown outside the strike zone to 
the batter's side. If we inverted the x-axis for the left-hand batters, 
we would get almost identical plots for the left-hand and right-hand batters.
Thus, I propose to invert (multiply by -1) the horizontal location of the ball
for the left-hand batters and use this inverted variable in the modeling. 
As for the vertical location of the ball, the effect is the same in each case. 
As long as the ball is not thrown below the strike zone, the higher the ball, 
the less likely the batter hits the ball into play. 


```{r fig.width=15, out.width="110%"}
## Creating variables for the modelling based on the vertical and horizontal
# location of the pitch


# Normalizing plate_x and plate_z for d1
# plate_x is 0 in the middle of the strike zone
# Let me normalize it so that -1 and 1 are side margins of strike zone
# The width of the strike zone is 17 inches or 17/12 feet
d1$plate_x_norm <- (d1$plate_x) / (17/12) * 2
# For the plate_z I want to normalize so that 0 is the bottom of the strike zone
# And 5.2 is the top. 5.2 because the height of the strike zone is on average
# 2.6 times bigger than the height
d1$plate_z_norm <- (d1$plate_z - d1$sz_bot) / (d1$sz_top - d1$sz_bot) * 5.2

d1_to_r <- d1 |> filter(stand == "R")
d1_to_l <- d1 |> filter(stand == "L")


k1 <- ggplot(d1_to_r, aes(x = plate_x_norm, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Location \n (right-hand batters only)", x = "Normalized Horizontal Ball Location", y = "Target") + 
  theme_grey(base_size = 16)

k2 <- ggplot(d1_to_l, aes(x = plate_x_norm, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Location \n (left-hand batters only)", x = "Normalized Horizontal Ball Location", y = "Target") + 
  theme_grey(base_size = 16)

k3 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = plate_z_norm, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Vertical Ball Location", x = "Normalized Vertical Ball Location", y = "Target") + 
  theme_grey(base_size = 16)

k1 + k2 + k3

# Conclustion based on visualization: 
# Invert the x axis if to lefties, no need to normalize
# Square the term and normalize by subtracting 1


# Creating needed variables and modelling based on the location of the ball
# Inverting horizontal location if the ball is thrown by left-hand player
d1$plate_x_norm_inv <- ifelse(d1$stand == "L", -d1$plate_x_norm, d1$plate_x_norm)
d1$plate_x_norm_inv_less_minus1 <- d1$plate_x_norm_inv * (d1$plate_x_norm_inv < -1)
d1$plate_z_norm_less_0 <- d1$plate_z_norm * (d1$plate_z_norm < 0)

# Normalizing to interpret the variable
# as the distance from the center of the plate in the modelling
# mean of plate_z_norm is about 3.45, 62% of the strike zone above lower edge. 
d1$plate_z_norm_norm <- d1$plate_z_norm - mean(d1$plate_z_norm)
```


The vertical movement of the ball is, on average, the same for 
the left-hand and right-hand batters. The horizontal movement is inverted ('mirrored') for
the left-hand batters, which is expected. 

```{r fig.width=13, fig.height = 7}
## Horizontal and vertical movements of the pitch

# Batter's p_throw matters obviously
# Batter's stand does not matter for the pitcher, however, the batter responds differently
# depending on the stand
# z is the same more or less
k1 <- ggplot(pitchers_data1, aes(x = avg_pfx_x_FF_L, y = avg_pfx_z_FF_L, color = p_throws)) +
  geom_point(alpha = 0.7) + 
  scale_colour_manual(values = c("red","blue")) + 
  labs(title = "Avg Pitch Movement to Left-Hand Batters \n (per pitcher)", 
       x = "Average Horizontal Ball Movement (per pitcher)", y = "Average Vertical Ball Movement (per pitcher)") +
  guides(color=guide_legend(title="Pitch Hand")) + 
   coord_fixed(ratio = 1.3, ylim = c(0.3, 1.7), xlim = c(-1.4, 1.4)) + 
  theme_grey(base_size = 14)

k2 <- ggplot(pitchers_data1, aes(x = avg_pfx_x_FF_R, y = avg_pfx_z_FF_R, color = p_throws)) +
  geom_point(alpha = 0.7) + 
  scale_colour_manual(values = c("red","blue")) + 
  labs(title = "Avg Pitch Movement to Right-Hand Batters \n (per pitcher)", 
       x = "Average Horizontal Ball Movement (per pitcher)", y = "Average Vertical Ball Movement (per pitcher)") + 
  guides(color=guide_legend(title="Pitch Hand")) + 
   coord_fixed(ratio = 1.3, ylim = c(0.3, 1.7), xlim = c(-1.4, 1.4)) + 
  theme_grey(base_size = 14)

k1 + k2
# 
# pitchers_data1 |> group_by(p_throws) |>
#   summarize(avg_pfx_x_FF_L = mean(avg_pfx_x_FF_L), 
#             avg_pfx_x_FF_R = mean(avg_pfx_x_FF_R),
#             avg_pfx_z_FF_L = mean(avg_pfx_z_FF_L), 
#             avg_pfx_z_FF_R = mean(avg_pfx_z_FF_R))


# Pfx's: logically, the horizonal movement of rhs and lhs pitchers are the opposite. 
# Vertical movement is the same and the movement does not matter on the stand 
# of the batter. 

# I still would like to look at l to r, r to l, l to l and r to r separately
# But in terms of variables I might consider the p_throws to interact with 
# pfx_x, but not with pfx_z, but we'll see
```

For the horizontal movement, the hand of the pitcher and the stand of the batter are the same, then the 
further the ball moves from the batter, the less likely he hits the ball into play. 
If the hand and stand are opposite, the effect is the opposite. Regardless of the hand and stand, the vertical movement has a stable negative association with the `TARGET`. 

```{r fig.width=12, out.width="80%"}
## Visulizations to come up with the most suitable variables for the modelling
# based on the horizontal and vertical movement of the ball. 


d1_l_to_r <- d1 |> filter(p_throws == "L", stand == "R")
d1_r_to_l <- d1 |> filter(p_throws == "R", stand == "L")
d1_l_to_l <- d1 |> filter(p_throws == "L", stand == "L")
d1_r_to_r <- d1 |> filter(p_throws == "R", stand == "R")

k1 <- ggplot(d1_l_to_r, aes(x = pfx_x, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Movement \n (left-hand pitch to right-hand only)", x = "Horizontal Ball Movement, feet", y = "Target") + 
  theme_grey(base_size = 16)

k2 <- ggplot(d1_l_to_l, aes(x = pfx_x, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Movement \n (left-hand pitch to left-hand only)", x = "Horizontal Ball Movement, feet", y = "Target") + 
  theme_grey(base_size = 16)

k3 <- ggplot(d1_r_to_r, aes(x = pfx_x, y = TARGET)) +
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Movement \n (right-hand pitch to right-hand only)", x = "Horizontal Ball Movement, feet", y = "Target") + 
  theme_grey(base_size = 16)

k4 <- ggplot(d1_r_to_l, aes(x = pfx_x, y = TARGET)) +
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Horizontal Ball Movement \n (right-hand pitch to left-hand only)", x = "Horizontal Ball Movement, feet", y = "Target") + 
  theme_grey(base_size = 16)

k1 + k2

k3 + k4


```

```{r out.width="40%"}
# An additional visualization to complement the above ones
ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = pfx_z, y = TARGET)) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Target vs Vertical Ball Movement", x = "Horizontal Ball Movement, feet", y = "Target") + 
  theme_grey(base_size = 16)
```


Like for the horizontal location of the ball, 
I created the inverted variable for horizontal movement (multiply by -1 if the stand is left). The relationship between this inverted variable and hit into play
for the same hand and stand and opposite hand and stand cases turns out to be similar. The slopes
are the same, so I only need to add the dummy variable indicating whether the stand and pitch are the same (or different). 

```{r out.width="50%"}
# Another plot for understanding 
d1$pfx_x_inv <- ifelse(d1$stand == "L", -d1$pfx_x, d1$pfx_x)
d1$diff_hand <- ifelse(d1$p_throws != d1$stand, 1, 0)

# So that's why it's just enought to just add the constant term diff hand
# pfx_x_inv has clear interpretation, so just state it in the report in the
# EDA part
ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = pfx_x_inv, y = TARGET, color = as.factor(diff_hand))) + 
  geom_jitter(height = 0.05, alpha = 0.2, size = 0.5) +
  labs(title = "Target vs Inverted Horizontal Ball Movement", x = "Inverted Horizontal Ball Movement, feet", y = "Target") +
  scale_color_manual(values = c("red", "blue")) +
  geom_smooth(method = 'lm') + 
  guides(color=guide_legend(title="Opposite Hands")) + 
  theme_grey(base_size = 16)

```


```{r}
## Creating variables based on the horizontal and vertical movement of the pitch
# for the modelling

d1$pfx_x_inv <- ifelse(d1$stand == "L", -d1$pfx_x, d1$pfx_x)
d1$diff_hand <- ifelse(d1$p_throws != d1$stand, 1, 0)

# Normalize for pfx_x_inv: Since I'm gonna interpret the model coefs at the same
# hand. Thus, I normalize at the mean for the same hand. 
# For pfx_z I just demean to interpret the model coefs at the mean
d1$pfx_x_inv_norm <- d1$pfx_x_inv - mean(d1$pfx_x_inv[d1$diff_hand == 0])
d1$pfx_z_norm <- d1$pfx_z - mean(d1$pfx_z)
```


The number of balls in at bat is positively associated with the `TARGET`, while
the association is negative for the number of strikes. As the number of balls increases, the pitchers become less risky, and at the same time, 
the batter does not rush and waits for an 'easy ball,' which results in a higher probability of getting hit into play. Linearity is observed for balls, 
while the unexpected jump at two strikes is observed for strikes. 
This might be because the batter becomes very risk-averse and tries to hit the ball with much less power. Otherwise, it is hard to explain
such phenomena.  

```{r fig.width=15, out.width="110%"}
## Score in at-bat: balls and strikes

d1_balls_strikes <- d1 |> group_by(balls, strikes) |> 
  summarize(avg_TARGET = mean(TARGET), n = n())
d1_balls_strikes$bs <- paste(d1_balls_strikes$balls, d1_balls_strikes$strikes, sep = "/")
  
# Score at bat
d1_balls <- d1 |> group_by(balls) |> 
  summarize(avg_TARGET = mean(TARGET), n = n())

d1_strikes <- d1 |> group_by(strikes) |> 
  summarize(avg_TARGET = mean(TARGET), n = n())



k1 <- ggplot(d1_balls, aes(x = balls, y = avg_TARGET)) +
  geom_bar(stat = "identity", fill = 'blue') +
  labs(title = "Average Target vs Balls", x = "Balls", y = "Average Target") + 
  theme_grey(base_size = 16)

# A little less intuitive why jumps at 2 strikes, 
# Implication of no foul at 2 strikes (but these fouls are recorded though)
# The batter becomes very risk-averse and tries just to hit the ball without 
# power
k2 <- ggplot(d1_strikes, aes(x = strikes, y = avg_TARGET)) +
  geom_bar(stat = "identity", fill = 'blue') +
  labs(title = "Average Target vs Strikes", x = "Strikes", y = "Average Target") + 
  theme_grey(base_size = 16)

k3 <- ggplot(d1_balls_strikes, aes(x = bs, y = avg_TARGET)) +
  geom_bar(stat = "identity", fill = 'blue') +
  labs(title = "Average Target vs Score At Bat", x = "Score At Bat, Balls/Strikes", y = "Average Target") + 
  theme_grey(base_size = 16)

k1 + k2 + k3

```

```{r}
## variables to consider

# I'll use the initial balls variable

# It turned out that this dummy variables (instead of number of strikes)
# has the better fit
d1$strikes2 <- ifelse(d1$strikes == 2, 1, 0)
```






### Appendix B: A couple of words about xWOBA as a measure of the quality of the hit. 

The main drawback of using the probability of hit into play as a measure of the 
batter's performance is that it treats any hit as a success. For example, home run and 
easy-to-catch pop-ups are treated similarly. We can use WOBA 
(weighted on-base average) weights to distinguish between these cases. wOBA
is the measure introduced by Tom Tango. In its 
[formula](https://library.fangraphs.com/offense/woba/), each outcome of the 
pitch (i.e. first base, home run) is multiplied by some coefficient that can be 
interpreted as the quality of the batter's performance at a certain pitch.
So for each pitch the corresponding wOBA weight can be attached depending on 
the outcome.

Such corresponding wOBA weight will depend not only on the batter's performance
(i.e., how good the ball was hit), but also on the performance of the fielders 
(i.e., the fielders can make an error and allow the batter to advance to the 
further base, which would result in a larger wOBA weight attached). 
To eliminate the effect of the fields, Statcast introduced
[Expected wOBA. aka xwOBA](https://www.mlb.com/glossary/statcast/expected-woba)
which mainly depends on the exit velocity and launch angle. As a result, xwOBA
can be interpreted as the "pure measure of the batter's performance at the pitch." 


I initially used xwOBA as a target variable in the analysis above. However, 
xwOBA turned out to have quite an unstable distribution. Most of the values in 
the data are zero, with some rare large values. This can be explained by the fact
that xwOBA has significant positive values only at selected exit 
velocity and launch angle values. Also, xwOBA is not quite a monotonic or smooth function 
of exit velocity and launch angle. As a result, the analysis above using 
xwOBA as the target variable resulted in unstable results, so I chose 
a less suitable but more stable probability of the hit instead. 

```{r out.width="50%"}
## xwOBA vs velocity angle

# only hits into play have non-zero xwOBA
hip_target <- d1[(d1$description == "hit_into_play") & (d1$launch_speed > 25), ]

p1 <- ggplot(hip_target, aes(x = launch_speed, y = launch_angle, color = xWOBA1)) +
  geom_point() +
  labs(title = "xwOBA vs exit velocity and launch angle", x = "exit velocity, mph", y = "launch angle, degrees") + 
  theme_grey(base_size = 16)+
  scale_color_gradient(low = "blue", high = "red", na.value = NA)
p1
# ggplotly(p1)
```

The only consistent (but not necessarily significant) findings are that xwOBA 
on average increases with the swing speed
(in general and for most players individually). In addition, 
xwOBA is almost always upper-bounded by 1 until the bat speed reaches a threshold 
in the range of 66 and 72 mph. xwOBA starts taking values more than 1 once this threshold 
is reached. The upper-bound becomes 2
(maximal possible value when it's very likely that the hit results in a home run) 
very quickly. It applies both to the aggregate and individual data.

```{r fig.width=8, out.width="50%"}
## Plotting relationship between xwOBA and swing speed in general and 
# for some players for illustration

g1 <- ggplot(d1, aes(x = bat_speed, y = xWOBA1)) +
  geom_jitter(height = 0.1, width = 0, alpha = 0.05) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "xwOBA vs swing speed, all players data", x = "swing speed, mph", y = "xwOBA") + 
  theme_grey(base_size = 16)

# 680776	Duran, Jarren, 606192	Hernández, Teoscar
batter0 <- 680776

batter_data <- d1 |> filter(batter == batter0)
batter_data_FF <- batter_data |> filter(pitch_type == "FF")


g2 <- ggplot(batter_data_FF, aes(x = bat_speed, y = xWOBA1)) +
  geom_jitter(height = 0.1, width = 0, alpha = 0.3) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "xwOBA vs swing speed, Jarren Duran", x = "swing speed, mph", y = "xwOBA") + 
  theme_grey(base_size = 16)

batter0 <- 606192


batter_data <- d1 |> filter(batter == batter0)
batter_data_FF <- batter_data |> filter(pitch_type == "FF")


g3 <- ggplot(batter_data_FF, aes(x = bat_speed, y = xWOBA1)) +
  geom_jitter(height = 0.1, width = 0, alpha = 0.3) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "xwOBA vs swing speed, Teoscar Hernandez", x = "swing speed, mph", y = "xwOBA") +  
  theme_grey(base_size = 16) 


g1
```

```{r fig.width=12, out.width="80%"}
g2 + g3
```


In further analysis, more data and more robust techniques can be used to
consider xwOBA as the target for the analysis above. 



### Appendix C: Exploratory Data Analysis, endogenous analyis.


```{r}
# Only 36% of the variation in the swing speed is explained by the pitcher. 
# If taking all pitches (not just FF like in this report), it would be even lower.

# model <- lm(bat_speed ~ as.factor(batter), data = d1)
# summary(model)$r.squared
# model <- 0
```



There is a negative relationship between normalized swing speed and pitch speed. This effect can be split into two parts. There is a slight negative effect on the pitcher's average pitch speed of the pitcher, and the negative effect on the pitch speed relative to the pitcher's average.
```{r fig.width = 15, out.width="110%"}
## A) Relationship with pitch speed

k1 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = release_speed, y = bat_speed_norm)) +
  geom_jitter(width = 0.1, alpha = 0.2, size = 0.5) +
  geom_smooth(method = 'lm', color = 'blue') +
  geom_smooth(color = 'red') + 
  labs(title = "Normalized Swing Speed vs Pitch Speed", 
       x = "Pitch speed, mph", y = "Normalized Swing Speed, mph") + 
  theme_grey(base_size = 16)

# Let's look at the pitch speed relative to the avg pitch speed of the pitcher
d1$release_speed_norm <- d1$release_speed - d1$avg_release_speed_FF

k2 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = release_speed_norm, y = bat_speed_norm)) +
  geom_jitter(width = 0, alpha = 0.2, size = 0.5) +
  geom_smooth(method = 'lm', color = 'blue') +
  geom_smooth(color = 'red') + 
  labs(title = "Normalized Swing Speed vs \n Normalized Pitch Speed", 
       x = "Normalized Pitch speed, mph", y = "Normalized Swing Speed, mph") + 
  theme_grey(base_size = 16)


k3 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = avg_release_speed_FF, y = bat_speed_norm)) +
  geom_jitter(alpha = 0.2, size = 0.5) +
  geom_smooth(method = 'lm', color = 'blue') +
  geom_smooth(color = 'red') + 
  labs(title = "Normalized Swing Speed vs \n Avg Pitch Speed (per pitcher)", 
       x = "Pitch speed, mph", y = "Normalized Swing Speed, mph") + 
  theme_grey(base_size = 16)

k1 + k2 + k3

```



Later innings tend to decrease the normalized swing speed gradually. The away team (bats during the top half-inning) starts to swing even slower than the home team in the later innings. 

```{r out.width="50%"}
## C) Inning number + score in runs (pre-pitch one) (+ top/bot)


# Inning: slighly affect, interesting that 1 is fast swing speed, 
# and then sharp slow down for 2 and the same swing speed for 3, 4, 5
# And gradually more for 6, 7, 8, 9
# After 9th more slowing down
# I can combine (2, 3, 4, 5), maybe (6, 7, 8) and 9+
d1$inning1 <- ifelse(d1$inning <= 9, d1$inning, '>9')
d1$inning2 <- ifelse(d1$inning <= 9, d1$inning, 10)



# Inning + top/bot: towards the end of the game (innings 7, 8, 9+), 
# the top (away) team starts to swing slower than the home team
# So I can create the variable top*inning7plus
d1g <- d1 |> group_by(inning2, inning_topbot) |> 
  summarize(mean_bat_speed = mean(bat_speed_norm), 
            n=n())
ggplot(d1g, aes(x = as.factor(inning2), y = as.factor(inning_topbot), fill = mean_bat_speed)) +
  geom_tile() +
  labs(title = "Normalized Average Swing Speed vs Half-Inning", 
       x = "Inning", y = "Half-Inning") + 
  guides(fill=guide_legend(title="mph")) + 
  scale_fill_gradient(high = 'red', low = 'blue') + 
  theme_grey(base_size = 16)



d1$inning3 <- ifelse(d1$inning1 %in% c("2", "3", "4", "5"), "2-5", d1$inning1)
d1$top_inning7plus <- ifelse(d1$inning2 >= 7 & d1$inning_topbot == "Top", 1, 0)

# Difference in runs is not significant to my surprise
```

Each base loaded has a relatively constant effect on the normalized swing speed, no matter how the other bases are loaded. Score at bat is one of the most significant variables. The number of balls is positively related to normalized swing speed, while the number of strikes is negatively related to normalized swing speed.

```{r fig.width=12, out.width="80%"}
## D) Bases loaded + score in at bat + outs when up (maybe score differential)

# Bases loaded matter: significant, and see that each base loaded give different effect
# Even using each combination gives better fit, but marginally, so I'll use 
# three dummies
d1$b1_loaded <- ifelse(!is.na(d1$on_1b), 1, 0)
d1$b2_loaded <- ifelse(!is.na(d1$on_2b), 1, 0)
d1$b3_loaded <- ifelse(!is.na(d1$on_3b), 1, 0)

d1$bases_loaded <- paste0(d1$b3_loaded, d1$b2_loaded, d1$b1_loaded)
d1$bases_loaded1 <- d1$b1_loaded + d1$b2_loaded + d1$b3_loaded

k1 <- ggplot(d1, aes(x = as.factor(bases_loaded), y = bat_speed_norm)) +
  geom_boxplot(fill = 'blue') + 
  labs(title = "Normalized Avg Swing Speed vs Bases Loaded", 
       x = "Bases Loaded, 3rd2nd1st", y = "Normalized Average Swing Speed, mph") + 
  theme_grey(base_size = 16)


# Outs: NOT SIGNIFICANT

# Score at bat: OMG, finally some really significant variable!!!
# Both number of strikes and balls relevant!!!
# You can either include score_at_bat or just balls and strikes
# Again, using combined score as factor gives only marginal improvement, 
# So I'll just use number of balls as strickes (as quantitative variables)
# Constant effect
d1$score_at_bat <- paste0(d1$balls, "/", d1$strikes)

k2 <- ggplot(d1, aes(x = as.factor(score_at_bat), y = bat_speed_norm)) +
  geom_boxplot(fill = 'blue') + 
  labs(title = "Normalized Avg Swing Speed vs \n Score at Bat", 
       x = "Score at Bat, Balls/Strikes", y = "Normalized Average Swing Speed, mph")  + 
  theme_grey(base_size = 16)

k1 + k2

# summary(lm(bat_speed_norm ~ as.factor(score_at_bat), data = d1))
# summary(lm(bat_speed_norm ~ balls + strikes, data = d1))
```


The ball's location is also one of the best predictors of the normalized swing speed. If the normalized horizontal location of the ball is inverted, then the relationship becomes the same in each of the 4 cases (right-hand pitcher throws to right-hand batter, l to l, r to l, and r to r), so there is no need to separate. The slope of the smooth regression line changes at 0. Similarly, the slope is changed at the value of 1.5 for the normalized vertical location. 


Remember that the normalized version of the location of the 
ball is derived such that the normalized horizontal location of the left edge of the strike zone (from the catcher's 
perspective) is equal to -1, and the right edge is equal to 1, and the 
the normalized vertical location of the bottom edge of the strike zone is equal to 0, and the
top edge is equal to 5.2 (this weird 5.2 value is because the height of the strike
zone is, on average, 2.6 higher than the width of the strike zone). 

```{r fig.width=12, out.width="80%"}
## E) Location of the ball (plate_x, plate_z) (check above) 
  # + quickly check the pfx_x, pfx_z i.e. ball movement down and to the side

# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# NOTICE THE DIFFERENT NUMBER OF ROWS FOR THE DFS BELOW, 
# MIGHT INDICATE DIFFERENT PROB OF FF DEPENDING ON HAND
# ALSO NOTICE FOR THE OTHER VARS ABOVE THAT ALSO SOME STUFF 
# INDICATES DIFFERENT PROB OF FF, SO THAT'S FOR FUTURE WORK


# 2 scenarios for vars creation: 
# 1) 4 vars: Just plate_z>2 and plate_z<2 for everyone, invert plate_x for 
# lefties and plate_x>0 and plate_x<0
# 2) 16 vars: 4 vars for each of the 4 cases above with thresholds specified
# for each case.

# 1st scenario
d1$plate_x_norm_inv <- ifelse(d1$stand == "L", -d1$plate_x_norm, d1$plate_x_norm)
d1$plate_x_norm_inv_less0 <- d1$plate_x_norm_inv * (d1$plate_x_norm_inv < 0)
d1$plate_z_norm_less1_5 <- d1$plate_z_norm * (d1$plate_z_norm < 1.5)

## The plots look the same in each scenario, so I show you the combined, smth like that
## Peaks at 0, change of slope
k1 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = plate_x_norm_inv, y = bat_speed_norm)) + 
  geom_jitter(height = 0, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Normalized Swing Speed vs \n Inverted Normalized Horizontal Ball Location", x = "Inverted Normalized Horizontal Ball Location", y = "Normalized Swing Speed, mph") + 
  theme_grey(base_size = 16)

## The plots look the same in each scenario, so I show you the combined, smth like that
## Change of slope at 1.5
k2 <- ggplot(d1[sample(1:nrow(d1), 30000, F), ], aes(x = plate_z_norm, y = bat_speed_norm)) + 
  geom_jitter(height = 0, alpha = 0.2, size = 0.5) +
  geom_smooth(color = 'red') +
  geom_smooth(method = 'lm', color = 'blue') +
  labs(title = "Normalized Swing Speed vs \n Normalized Vertical Ball Location", x = "Normalized Vertical Ball Location", y = "Normalized Swing Speed, mph") + 
  theme_grey(base_size = 16)

k1 + k2




# # 1st scenario: is enough!!!
# summary(lm(bat_speed_norm ~ release_speed + avg_release_speed_FF + 
#              plate_x_norm_inv + plate_x_norm_inv_less0 + 
#              plate_z_norm + plate_z_norm_less2 + pfx_x:stand + pfx_z + 
#              score_at_bat, data = d1))


# # adding the average plate_x, plate_z of the pitcher
# # Significant, I'll keep them: UPD NOT HERE, I THINK THEY ARE SIGNIFICANT
# # BUT DO NOT AFFECT THE FIT, SO I DON'T WORRY AT THIS POINT
# d1$avg_plate_x_FF_L_standL <- ifelse(d1$stand == "L", d1$avg_plate_x_FF_L, 0)
# d1$avg_plate_x_FF_L_standL_inv <- -d1$avg_plate_x_FF_L_standL
# d1$avg_plate_x_FF_R_standR <- ifelse(d1$stand == "R", d1$avg_plate_x_FF_R, 0)
# summary(lm(bat_speed_norm ~ release_speed + avg_release_speed_FF + 
#              plate_x_norm_inv_more0 + plate_x_norm_inv_less0 + 
#              plate_z_norm_more2 + plate_z_norm_less2 + 
#              avg_plate_x_FF_L_standL + avg_plate_x_FF_R_standR + 
#              avg_plate_z_FF + 
#              pfx_x:stand + pfx_z + 
#              score_at_bat, data = d1))

```

In all four cases, horizontal and vertical movement did not seem significant in bivariate analysis. However, these variables showed significance in the modeling. 

```{r}
## Both pfx_x and pfx_z did not seem to be significant on bivariate analysis 
# no matter how I was splitting the data. However, in the final model 
# the variables turned out to be significant. 

# Well, no clear dependence, bigger dip down is associated with 
# slightly higher swing speed, and this association might be slightly different
# for lh and rh batters, so the only I can propose is to split the pfx_z 
# depending on the stand of the batter
```

Similarly, the location of the bottom of the strike zone (normalized relative to the average location of the bottom of the batter's strike zone) is significant. I thought this variable would be significant because I thought that how the batter bends his legs might affect the normalized swing speed, which can be proxied by the vertical coordinates of the top and bottom of the strike zone. The location of the top of the strike zone (normalized) is not significant, though. 
```{r}
## E1) sz_top and sz_bottom, maybe swing speed mechanically depends on the position of the player!!!

d1 <- d1 |> group_by(batter, stand) |>
  mutate(mean_sz_bot = mean(sz_bot),
         mean_sz_top = mean(sz_top))
d1$sz_top_demean <- d1$sz_top - d1$mean_sz_top
d1$sz_bot_demean <- d1$sz_bot - d1$mean_sz_bot
# 
# ggplot(d1, aes(x = sz_bot_demean, y = bat_speed_norm)) +
#   geom_point(alpha = 0.05, size = 0.1) +
#   geom_smooth(method = 'lm') +
#   labs(title = "Relationship between swing speed and the bottom of the strike zone", 
#        x = "Bottom of the strike zone", y = "Swing speed normalized") + 
#   geom_smooth(color = 'blue') + 
#   coord_cartesian(xlim = c(-0.1, 0.1))
```






